---
output: html_document
---
<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{rsdmx quickstart guide}
-->

# rsdmx quickstart guide

The goal of this document is to get you up and running with rsdmx as quickly as possible.

``rsdmx`` provides a set of classes and methods to read data and metadata documents exchanged through the Statistical Data and Metadata Exchange (SDMX) framework. The package currently focuses on the SDMX XML standard format (SDMX-ML).

For more information about SDMX, you can visit the [SDMX website](http://sdmx.org/). 

[rsdmx](http://cran.r-project.org/web/packages/rsdmx/index.html) offers a low-level set of tools to read **data** and **metadata** in SDMX format. Its strategy is to make it very easy for the user. For this, a unique function named ``readSDMX`` has to be used, whatever it is a ``data`` or ``metadata`` document, or if it is ``local`` or ``remote`` datasource.


## Install rsdmx

``rsdmx`` can be installed from CRAN or from its development repository hosted in Github (using the ``devtools`` package):

```{r, echo = FALSE}
devtools::install_github("opensdmx/rsdmx")
```

## Load rsdmx

```{r, echo = FALSE}
library(rsdmx)
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

To load rsdmx in R, do the following:

```{r, echo = FALSE}
library(rsdmx)
```

## Read dataset documents

This section will introduce you on how to read SDMX *dataset* documents, either from remote datasources, or from local SDMX files.

##### Read _remote_ datasets

The following code snipet shows you how to read a dataset from a remote data source, taking as example the [FAO data portal](http://data.fao.org/sdmx/index.html): [http://data.fao.org/sdmx/repository/data/CROP_PRODUCTION/.156.5312../FAO?startPeriod=2008&endPeriod=2008](http://data.fao.org/sdmx/repository/data/CROP_PRODUCTION/.156.5312../FAO?startPeriod=2008&endPeriod=2008)

```{r, echo = FALSE}
myUrl <- "http://data.fao.org/sdmx/repository/data/CROP_PRODUCTION/.156.5312../FAO?startPeriod=2008&endPeriod=2008"
dataset <- readSDMX(myUrl)
stats <- as.data.frame(dataset) 
```

You can try it out with other datasources, such as:
* [**OECD StatExtracts portal**](http://stats.oecd.org): [http://stats.oecd.org/restsdmx/sdmx.ashx/GetData/MIG/TOT../OECD?startTime=2000&endTime=2011](http://stats.oecd.org/restsdmx/sdmx.ashx/GetData/MIG/TOT../OECD?startTime=2000&endTime=2011)
* [**EUROSTAT portal**](http://epp.eurostat.ec.europa.eu/portal/page/portal/sdmx_web_services/getting_started/rest_sdmx_2.1): [http://ec.europa.eu/eurostat/SDMX/diss-web/rest/data/cdh_e_fos/..PC.FOS1.BE/?startperiod=2005&endPeriod=2011](http://ec.europa.eu/eurostat/SDMX/diss-web/rest/data/cdh_e_fos/..PC.FOS1.BE/?startperiod=2005&endPeriod=2011)
* [**European Central Bank (ECB)**](https://sdw-wsrest.ecb.europa.eu): [https://sdw-wsrest.ecb.europa.eu/service/data/DD/M.SE.BSI_STF.RO.4F_N](https://sdw-wsrest.ecb.europa.eu/service/data/DD/M.SE.BSI_STF.RO.4F_N)

The online rsdmx documentation also provides a list of data providers, either from international or national institutions.


##### Read _local_ datasets

This example shows you how to use ``rsdmx`` with _local_ SDMX files, previously downloaded from [EUROSTAT](http://ec.europa.eu/eurostat).

```{r, echo = FALSE}
#bulk download from Eurostat
tf <- tempfile(tmpdir = tdir <- tempdir()) #temp file and folder
download.file("http://ec.europa.eu/eurostat/estat-navtree-portlet-prod/BulkDownloadListing?sort=1&file=data%2Frd_e_gerdsc.sdmx.zip", tf)
sdmx_files <- unzip(tf, exdir = tdir)

sdmx <- readSDMX(sdmx_files[2], isURL = FALSE)
stats <- as.data.frame(sdmx)
head(stats)

```

By default, ``readSDMX`` considers the data source is remote. To read a local file, add ``isURL = FALSE``.

## Read metadata documents

This section will introduce you on how to read SDMX **metadata** documents, including ``concepts``, ``codelists`` and complete ``data structure definitions`` (DSD)


### Concepts

Read concept schemes from [FAO data portal](http://data.fao.org/sdmx/index.html)

```{r, echo = FALSE}
csUrl <- "http://data.fao.org/sdmx/registry/conceptscheme/FAO/ALL/LATEST/?detail=full&references=none&version=2.1"
csobj <- readSDMX(csUrl)
csdf <- as.data.frame(csobj)
head(csdf)
```

### Codelists

Read codelists from [FAO data portal](http://data.fao.org/sdmx/index.html)
```{r, echo = FALSE}
clUrl <- "http://data.fao.org/sdmx/registry/codelist/FAO/CL_FAO_MAJOR_AREA/0.1"
clobj <- readSDMX(clUrl)
cldf <- as.data.frame(clobj)
head(cldf)
```

###Data Structure Definition (DSD)

This example illustrates how to read a complete DSD using a [OECD StatExtracts portal](http://stats.oecd.org) data source.

```{r, echo = FALSE}
dsdUrl <- "http://stats.oecd.org/restsdmx/sdmx.ashx/GetDataStructure/TABLE1"
dsd <- readSDMX(dsdUrl)
```

``rsdmx`` is implemented in object-oriented way with ``S4`` classes and methods. The properties of ``S4`` objects are named ``slots`` and can be accessed with the ``slot`` method:

The following code snippet allows to extract the list of ``codelists`` contained in the DSD document, and read one codelist as ``data.frame``.

```{r, echo = FALSE}
#get codelists from DSD
cls <- slot(dsd, "codelists")
codelists <- sapply(slot(cls, "codelists"), function(x) slot(x, "id")) #get list of codelists
codelist <- as.data.frame(slot(dsd, "codelists"), codelistId = "CL_TABLE1_FLOWS") #get a codelist
```

In a similar way, the ``concepts`` of the dataset can be extracted from the DSD and read as ``data.frame``.

```{r, echo = FALSE}
#get concepts from DSD
concepts <- as.data.frame(slot(dsd, "concepts"))
```
