<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{rsdmx quickstart guide}
-->

# rsdmx quickstart guide

The goal of this document is to get you up and running with rsdmx as quickly as possible.

``rsdmx`` provides a set of classes and methods to read data and metadata documents exchanged through the Statistical Data and Metadata Exchange (SDMX) framework.

## SDMX - a short introduction

The SDMX framework provides two sets of standard specifications to facilitate the exchange of statistical data:

* standard formats
* web-service specifications

SDMX allows to disseminate both **data** (a dataset) and **metadata** (the description of the dataset).

For this, the SDMX standard provides various types of _documents_, also known as _messages_. Hence there will be:

* **data** SDMX-ML _documents_. The two main _document_ types are the ``Generic`` and ``Compact`` ones. The latter aims to provide a more compact XML document. They are other data _document_ derivating from the main types.
* **metadata** SDMX-ML _documents_. The main metadata _document_ is known a ``Data Structure Definition`` (DSD). As its name indicates, it _describes_ the structure and organization of a dataset, and will generally include all the master/reference data used to characterize a dataset. The 2 main types of metadata are (1) the ``concepts``, which correspond the _dimensions_ and/or _attributes_ of the dataset, and (2) the ``codelists`` which inventory the possible values to be used in the representation of _dimensions_ and _attributes_.

For more information about the SDMX standards, you can visit the [SDMX website](http://sdmx.org/), or this [introduction by EUROSTAT](https://webgate.ec.europa.eu/fpfis/mwikis/sdmx/index.php/SDMX).

## How to deal with SDMX in R

[rsdmx](http://cran.r-project.org/web/packages/rsdmx/index.html) offers a low-level set of tools to read **data** and **metadata** in the SDMX-ML format. Its strategy is to make it very easy for the user. For this, a unique function named ``readSDMX`` has to be used, whatever it is a ``data`` or ``metadata`` document, or if it is ``local`` or ``remote`` datasource.

* what ``rsdmx`` does support: a SDMX format abstraction library, with focus on the the main SDMX standard XML format (SDMX-ML), and the support of the three format standard versions (``1.0``, ``2.0``, ``2.1``)
* what ``rsdmx`` does NOT support: an interface to SDMX web-services. Although in a near future ``rsdmx`` plans to support ``helpers`` to build SDMX data queries according to SDMX web-services, its primary strategy is to give an emphasis on the SDMX-ML **format**. As result, ``rsdmx`` can be applied to a wider range of datasources, that do not necessarily provide SDMX standard compliant web-services or even may require to download the data prior its reading in R. 

Let's see then how to use ``rsdmx``!

## Install rsdmx

``rsdmx`` can be installed from CRAN or from its development repository hosted in Github (using the ``devtools`` package). For the latter, you will need the ``devtools`` package and run:

```{r, eval=FALSE}
devtools::install_github("opensdmx/rsdmx")
```

## Load rsdmx

To load rsdmx in R, do the following:

```{r, eval=FALSE}
library(rsdmx)
```

## Read dataset documents

This section will introduce you on how to read SDMX *dataset* documents, either from remote datasources, or from local SDMX files.

### Read _remote_ datasets

The following code snipet shows you how to read a dataset from a remote data source, taking as example the [FAO data portal](http://data.fao.org/sdmx/index.html): [http://data.fao.org/sdmx/repository/data/CROP_PRODUCTION/.156.5312../FAO?startPeriod=2008&endPeriod=2008](http://data.fao.org/sdmx/repository/data/CROP_PRODUCTION/.156.5312../FAO?startPeriod=2008&endPeriod=2008)

```{r, eval=FALSE}
myUrl <- "http://data.fao.org/sdmx/repository/data/CROP_PRODUCTION/.156.5312../FAO?startPeriod=2008&endPeriod=2008"
dataset <- readSDMX(myUrl)
stats <- as.data.frame(dataset) 
```

You can try it out with other datasources, such as:

* [**OECD StatExtracts portal**](http://stats.oecd.org): [http://stats.oecd.org/restsdmx/sdmx.ashx/GetData/MIG/TOT../OECD?startTime=2000&endTime=2011](http://stats.oecd.org/restsdmx/sdmx.ashx/GetData/MIG/TOT../OECD?startTime=2000&endTime=2011)
* [**EUROSTAT portal**](http://epp.eurostat.ec.europa.eu/portal/page/portal/sdmx_web_services/getting_started/rest_sdmx_2.1): [http://ec.europa.eu/eurostat/SDMX/diss-web/rest/data/cdh_e_fos/..PC.FOS1.BE/?startperiod=2005&endPeriod=2011](http://ec.europa.eu/eurostat/SDMX/diss-web/rest/data/cdh_e_fos/..PC.FOS1.BE/?startperiod=2005&endPeriod=2011)
* [**European Central Bank (ECB)**](https://sdw-wsrest.ecb.europa.eu): [https://sdw-wsrest.ecb.europa.eu/service/data/DD/M.SE.BSI_STF.RO.4F_N](https://sdw-wsrest.ecb.europa.eu/service/data/DD/M.SE.BSI_STF.RO.4F_N)

The online rsdmx documentation also provides a list of data providers, either from international or national institutions.

### Read _local_ datasets

This example shows you how to use ``rsdmx`` with _local_ SDMX files, previously downloaded from [EUROSTAT](http://ec.europa.eu/eurostat).

```{r, eval=FALSE}
#bulk download from Eurostat
tf <- tempfile(tmpdir = tdir <- tempdir()) #temp file and folder
download.file("http://ec.europa.eu/eurostat/estat-navtree-portlet-prod/BulkDownloadListing?sort=1&file=data%2Frd_e_gerdsc.sdmx.zip", tf)
sdmx_files <- unzip(tf, exdir = tdir)

sdmx <- readSDMX(sdmx_files[2], isURL = FALSE)
stats <- as.data.frame(sdmx)

```

By default, ``readSDMX`` considers the data source is remote. To read a local file, add ``isURL = FALSE``.

## Read metadata documents

This section will introduce you on how to read SDMX **metadata** documents, including ``concepts``, ``codelists`` and complete ``data structure definitions`` (DSD)


### Concepts

Read concept schemes from [FAO data portal](http://data.fao.org/sdmx/index.html)

```{r, eval=FALSE}
csUrl <- "http://data.fao.org/sdmx/registry/conceptscheme/FAO/ALL/LATEST/?detail=full&references=none&version=2.1"
csobj <- readSDMX(csUrl)
csdf <- as.data.frame(csobj)
```

### Codelists

Read codelists from [FAO data portal](http://data.fao.org/sdmx/index.html)
```{r, eval=FALSE}
clUrl <- "http://data.fao.org/sdmx/registry/codelist/FAO/CL_FAO_MAJOR_AREA/0.1"
clobj <- readSDMX(clUrl)
cldf <- as.data.frame(clobj)
```

###Data Structure Definition (DSD)

This example illustrates how to read a complete DSD using a [OECD StatExtracts portal](http://stats.oecd.org) data source.

```{r, eval=FALSE}
dsdUrl <- "http://stats.oecd.org/restsdmx/sdmx.ashx/GetDataStructure/TABLE1"
dsd <- readSDMX(dsdUrl)
```

``rsdmx`` is implemented in object-oriented way with ``S4`` classes and methods. The properties of ``S4`` objects are named ``slots`` and can be accessed with the ``slot`` method. The following code snippet allows to extract the list of ``codelists`` contained in the DSD document, and read one codelist as ``data.frame``.

```{r, eval=FALSE}
#get codelists from DSD
cls <- slot(dsd, "codelists")
codelists <- sapply(slot(cls, "codelists"), function(x) slot(x, "id")) #get list of codelists
codelist <- as.data.frame(slot(dsd, "codelists"), codelistId = "CL_TABLE1_FLOWS") #get a codelist
```

In a similar way, the ``concepts`` of the dataset can be extracted from the DSD and read as ``data.frame``.

```{r, eval=FALSE}
#get concepts from DSD
concepts <- as.data.frame(slot(dsd, "concepts"))
```
